Natural transformations

### 2017-10-11 papers/base-3/paper-3.tex
Update section on functor category

\section*{Functor Category}
\begin{defn}[Functor Category]
  Natural transformations, as other types of maps that we have encountered earlier
  can be composed. That is, for any two categories $\cat{A}$ and $\cat{B}$,
  functors $F, G, H$ between them, natural transformations $\alpha \from F \to
  G$ and $\beta \from G \to H$:
  \vspace{-2ex}
  \[
  \xymatrix@C+1.5em{
  \cat{A}
  \ruppertwocell<8>^F{\alpha}
  \ar[r]|G
  \rlowertwocell<-8>_H{\beta}
  &
  \cat{B}
  },
  \]
  \vspace{-3ex}\\
  there is a composite natural transformation $\beta \of \alpha$
  \[
  \xymatrix@C+1.5em{
  \cat{A} \rtwocell^F_H{\hspace{.8em}\beta\of\alpha} &\cat{B}
  }
  \]
  defined by $(\beta\of\alpha)_A = \beta_A \of \alpha_A$ for all $A \in \cat{A}$.
  There is also an identity natural transformation
  \[
  \xymatrix@C+1em{
  \cat{A} \rtwocell^F_F{\hspace{.5em}1_F} &\cat{B}
  }
  \]
  on any functor $F$, defined by $(1_F)_A = 1_{F(A)}$.

  This gives rives to a category whose objects are the functors from $\cat{A}$ to
  $\cat{B}$ and whose maps are the natural transformations between them. This is
  called the \demph{functor category} from $\cat{A}$ to $\cat{B}$, and written as
  $\ftrcat{\cat{A}}{\cat{B}}$ or $\cat{B}^\cat{A}$.
\end{defn}

\begin{exmp}
  Let $2$ be the discrete category with two objects. A functor from $2$ to a
  category $\cat{B}$ is a pair of objects of $\cat{B}$, and a natural
  transformation is a pair of maps.  The functor category $\ftrcat{2}{\cat{B}}$
  is therefore isomorphic to the product category $\cat{B} \times \cat{B}$.
  This fits well with the alternative notation $\cat{B}^2$ for the functor
  category.
\end{exmp}

\begin{exmp}
  Take ordered sets $A$ and $B$, viewed as categories. Given order-preserving
  maps $\parpairi{A}{B}{f}{g}$, viewed as functors, there is at most one
  natural transformation
  \[
  \xymatrix{
  A \rtwocell^f_g &B,
  }
  \]
  and there is one if and only if $f(a) \leq g(a)$ for all $a \in A$.  (The
  naturality axiom~\eqref{eq:nat} holds automatically, because in an ordered
  set, all diagrams commute.)  So $\ftrcat{A}{B}$ is an ordered set too; its
  elements are the order-preserving maps from $A$ to $B$, and $f \leq g$ if and
  only if $f(a) \leq g(a)$ for all $a \in A$.
\end{exmp}

\begin{defn}
  Let $\cat{A}$ and $\cat{B}$ be categories. A \demph{natural isomorphism}
  between functors from $\cat{A}$ to $\cat{B}$ is an isomorphism in
  $\ftrcat{\cat{A}}{\cat{B}}$.
\end{defn}

@@

Then, we say that functors $F$ and $G$ are \demph{naturally isomorphic} if
there is a natural isomorphism from $F$ to $G$. Note that this natural
isomorphism is just isomorphism in $\ftrcat{\cat{A}}{\cat{B}}$, and we already
have notation for this: $F \iso G$.

\begin{defn}
  Given functors $\parpair{\cat{A}}{\cat{B}}{F}{G}$, we say that
  \[
  F(A) \iso G(A) \text{ \demph{naturally in} } A
  \]
  if $F$ and $G$ are naturally isomorphic.
\end{defn}

This alternative terminology can be understood as follows.  If $F(A) \iso
G(A)$ naturally in $A$ then certainly $F(A) \iso G(A)$ for each individual
$A$, but more is true: we can choose isomorphisms $\alpha_A\from F(A) \to
G(A)$ in such a way that the naturality axiom~\eqref{eq:nat} is satisfied.

\begin{exmp}
  Let $F, G\from \cat{A} \to \cat{B}$ be functors from a discrete category
  $\cat{A}$ to a category $\cat{B}$.  Then $F \iso G$ if and only if $F(A) \iso
  G(A)$ for all $A \in \cat{A}$.

  So in \emph{this} case, $F(A) \iso G(A)$ naturally in $A$ if and only if
  $F(A) \iso G(A)$ for all $A$.  But this is only true because $\cat{A}$ is
  discrete.  In general, it is emphatically false.  There are many examples of
  categories and functors $\parpairi{\cat{A}}{\cat{B}}{F}{G}$ such that $F(A)
  \iso G(A)$ for all $A \in \cat{A}$, but not \emph{naturally} in $A$.
\end{exmp}

>>>>

\section*{Functor Category}

Consider a composition $\beta \of \alpha$ between some natural transformations
$\alpha \from F \to G$ and $\beta \from G \to H$.
\[
  \xymatrix@C+1.5em{
    \cat{A}
\ruppertwocell<8>^F{\alpha}
\ar[r]|G
\rlowertwocell<-8>_H{\beta}
&
\cat{B}
},
\]
\vspace{-3ex}\\
We can define such composite natural transformation
\[
  \xymatrix@C+1.5em{
    \cat{A} \rtwocell^F_H{\hspace{.8em}\beta\of\alpha} &\cat{B}
  }
\]
with $(\beta\of\alpha)_A = \beta_A \of \alpha_A$ for all $A \in \cat{A}$.
(Recall that $\beta_A$ and $\alpha_A$ are moprhisms in $\cat{B}$.)

We can also define an identity natural transformation
\[
  \xymatrix@C+1em{
    \cat{A} \rtwocell^F_F{\hspace{.5em}1_F} &\cat{B}
  }
\]
on any functor $F$ with $(1_F)_A = 1_{F(A)}$.

Thus we have a definition of a category:

\begin{defn}[Functor Category]
  The \demph{functor category} from a category $\cat{A}$ to a category
  $\cat{B}$ (written as $\ftrcat{\cat{A}}{\cat{B}}$ or $\cat{B}^\cat{A}$) is a
  category whose objects are the functors from $\cat{A}$ to $\cat{B}$. The maps
  in this category are the natural transformations between functors.
\end{defn}

\begin{exmp}[Functors from a discrete category]
  Consider the discrete category with two objects, $2$. Any functor from $2$ to
  some another category $\cat{B}$ will be a pair of objects in $\cat{B}$. A
  natural transformation between such functors then is a pair of maps between
  these objects.  Note that the functor category $\ftrcat{2}{\cat{B}}$ is
  isomorphic to the product category $\cat{B} \times \cat{B}$ and the
  superscript notation for the functor category $\cat{B}^2$ reflects this
  nicely.
\end{exmp}

\begin{exmp}[Maps in ordered sets]
  Consider ordered sets $A$ and $B$, viewed as categories. The maps
  $\parpairi{A}{B}{f}{g}$ that preserve order can be then considered as
  functors. Note that there is at most one natural transformation between such
  functors and it exists if and only if $f(a) \leq g(a)$ for all $a \in A$.  In
  this case we have the naturality axiom~\eqref{eq:nat} automatically, since
  all diagrams commute in an ordered set. Moreover, the functor category
  $\ftrcat{A}{B}$ is an ordered set itself, where elements are the
  order-preserving maps from $A$ to $B$, and there a morphism $f \leq g$
  between $f$ and $g$ if and only if $f(a) \leq g(a)$ for all $a \in A$.
\end{exmp}

\begin{defn}[Natural isomorphism]
  Let $\cat{A}$ and $\cat{B}$ be categories. A \demph{natural isomorphism}
  between functors from $\cat{A}$ to $\cat{B}$ is an isomorphism in
  $\ftrcat{\cat{A}}{\cat{B}}$.
\end{defn}

@@

\begin{defn}[Isomorphic functors]
  We call functors $F$ and $G$ \demph{naturally isomorphic} if there is a
  natural isomorphism from $F$ to $G$, which is just an isomorphism in
  $\ftrcat{\cat{A}}{\cat{B}}$. The notation for isomorphic functors is the same
  as for isomorphic objects: $F \iso G$.
\end{defn}

\begin{defn}[Isomorphic naturally in an object]
  Given functors $\parpair{\cat{A}}{\cat{B}}{F}{G}$, we say that
  \[
  F(A) \iso G(A) \text{ \demph{naturally in} } A
  \]
  if $F$ and $G$ are naturally isomorphic.
\end{defn}

Note that the definition says that we not only have $F(A) \iso G(A)$ for each
individual $A$, but also there is an isomorphism $\alpha_A\from F(A) \to G(A)$
that satisfies the naturality axiom~\eqref{eq:nat}.

\begin{exmp}
  Consider two functors $F, G\from \cat{A} \to \cat{B}$ where $\cat{A}$ is a
  discrete category. We have $F \iso G$ if and only if $F(A) \iso G(A)$ for all
  $A \in \cat{A}$. If it is the case, then we also have that $F(A) \iso G(A)$
  naturally in $A$ for all $A$.

  Be aware that it is not always so: there are examples of categories and
  functors $\parpairi{A}{B}{F}{G}$ where $F(A) \iso G(A)$ for all $A \in
  \cat{A}$, but not naturally in $\cat{A}$. One of the examples arises if one
  considers two functors that take the objects of the category of finite sets
  and bijections to the sets of permutations and the sets of total orders
  correspondingly.  There one can find isomorphisms between the sets of
  permutations and the sets of total orders, but no natural isomorphism between
  such functors.
\end{exmp}
